---
title: "Alcohol consumption and the risks of adult diseases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alcohol disease risks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: disease-risks.bib
link-citations: yes
citation_package: natbib
biblio-style: vancouver
urlcolor: blue
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos = 'H'
)
```

```{r setup, include = FALSE, results = 'hide', warning = FALSE, eval = T}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(testthat))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tobalcepi))
suppressPackageStartupMessages(library(readxl))
```

# Background
The latest versions of the dose-response curve linking alcohol consumption to disease are available in a [short report](https://figshare.shef.ac.uk/articles/Alcohol-attributable_diseases_and_dose-response_curves_for_the_Sheffield_Alcohol_Policy_Model_version_4_0/6819689/2) [@Angus2018]. This vignette does not aim to reproduce that report - instead this vignette explains how the risk functions within the `tobalcepi` code can be accessed and used.   


# Disease list



# Partially-attributable chronic conditions




# Partially-attributable acute conditions




# Wholly-attributable acute conditions





# Wholly-attributable acute conditions



# Use of relative risks in SAPM




# References




































































































